Emergent dual scaling of riverine biodiversity

### 2021-03-23 manuscript_pnas.Rmd
Create manuscript_pnas.Rmd

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
output:
  pdf_document:
    latex_engine: xelatex
csl: science.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

**Abstract:** Larger ecosystems support more species; this ubiquitous pattern is the foundation of current conservation schemes. However, many ecosystems possess a complex spatial structure that cannot be represented by area, and the role of such complexity in controlling biodiversity is largely unknown. Here, we use theory and extensive fish community data from two distinct geographic regions (Japan and United States) to show that ecosystem size and complexity dictate riverine biodiversity. We found that larger and more branched 'complex' river networks harbored greater species richness due to increased space and environmental heterogeneity. The complexity effect was comparable to the size effect, and this pattern has emerged regardless of ecological contexts. The dual control of biodiversity may be a pervasive feature that has far-reaching implications for biodiversity conservation.

@@

Using theory and extensive community data, our study provides the strong evidence that ecosystem size and complexity play equally important roles in shaping riverine biodiversity. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

>>>>

---
title: "Emergent dual scaling of riverine biodiversity"
output:
  pdf_document:
    latex_engine: xelatex
csl: pnas.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

**Abstract:** A prevailing thought suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions (Hokkaido Island in Japan and the midwestern United States) to show that ecosystem size and complexity play equally-important roles in scaling biodiversity in rivers. The recurrent merging of various streams and rivers forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. Both relationships were linear on a log-log scale, suggesting the general power-law scaling along the two orthogonal dimensions of ecosystem structure. In support of this theoretical prediction, the robust dual power-law scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fauna with distinct evolutionary histories. The dual scaling of biodiversity may be a pervasive emergent property of branching networks with important implications for biodiversity conservation.

**Significance** Larger ecosystems support more species; this ubiquitous pattern is the foundation of current conservation schemes. However, many ecosystems possess a complex spatial structure that cannot be represented by area, and the role of such complexity in controlling biodiversity is largely unknown. Here, we use theory and extensive fish community data to show that ecosystem size and complexity dictate riverine biodiversity. We found that larger and more branched 'complex' river networks harbored greater species richness due to increased space and environmental heterogeneity. The complexity effect was comparable to the size effect, and this pattern has emerged regardless of ecological contexts. The dual scaling of biodiversity may be a pervasive feature that has far-reaching implications for biodiversity conservation.

**Key words:** metacommunity, species diversity, scaling law, network theory, freshwater

@@

Using theory and extensive community data, our study provides strong evidence that ecosystem size and complexity play equally important roles in shaping riverine biodiversity. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].
